National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-27 Manuscript_Bayesian.Rmd
close to final revision - ben to proof on overleaf.

---
title             : "Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

@@

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5415 words in text body, 1155 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
# header-includes:
#    - \usepackage{tcolorbox}
#    - \usepackage{paralist}
#    - \usepackage{multicol}
#    - \usepackage{graphics}
#    - \usepackage{rotating}
#    - \usepackage{longtable}
#    - \usepackage[amsmath]
---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```

# ```{r runanalysis, include = FALSE}
# source('short.R')
# ```
The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three credible challenges.

@@

@@

## Statistical Analyses

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years. The model equations are given below and priors are given below.

## Potential Confounders and Missing Data

Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves. In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. There are increasing worries

@@

### System and packages

The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"

## Code

All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
```

@@

\begin{table}
\begin{center}
\scalebox{0.5}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00; 0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$          \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01; 0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$           \\
SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS}
\end{center}
\end{table}

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
```

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
```

@@

### Authors contributions
BH conceived of the study. BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB. EW improved the approach with BH and JB and revised the manuscript. DD revised the manuscript. CS was responsible for NZAVS data collection and revised the manuscript. JB improved the approach with BH and EW, performed a data re-analysis, and revised the manuscript.

@@

# Appendix

\newpage
## Model Equations

### Priors for the Personal Wellbeing model

>>>>

---
title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

@@

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "4685 words in text body, 1191 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]

---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```
```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(#fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                     # fig.width= 10,
                     # fig.height=8,
                      tidy = "styler")
```

The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three credible challenges.

@@

Table 1 about here

\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{0.65}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule

\end{tabular}
}
\end{table}

@@

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years. The model equations are given below and priors are given below.

## Potential Confounders and Missing Data 
Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves. In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. There are increasing worries

@@

## System and packages
The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"

## Code
All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).

@@

@@

\begin{table}
\begin{center}
\scalebox{0.8}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00; 0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$          \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01; 0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$           \\
SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS}
\end{center}
\end{table}

Figure 1 about here
```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefficientplotsbayes.png")
```

## Personal Well-Being
We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).

Within this model:

  - The effect of years is positive is reliable (posterior median = 0.025, HDI [0.020, 0.031] pd = 100.00%).  We infer that personal well-being increased in the skeptic population between 2010 and 2018.
  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median = -0.058, HDI [-0.080, -0.035], pd = 100.00%). We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
  - The effect of beliefs in *God And Spirit*  is reliably negative (posterior median = -0.064, HDI [-0.091, -0.037], pd = 100.00%). We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.002 HDI [-0.039,  0.033], pd = 53.50%). We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.

## Personal Well-Being: years X beliefs

```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
```

The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.010, HDI [ 0.003,  0.017], pd = 99.02%). We infer that people with beliefs in a spirit or life force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.008, 0.021], pd = 99.98%). We infer that people who believe in a spirit or life force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.002, HDI [-0.009,  0.015], pd =  63.73%).  We infer that people who believe in a God but not a spirit or life force did not grow at a faster rate in personal well-being than skeptics grew.

## Life Satisfaction

We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see model equations, below).

Within this model:

  - The effect of years is positive is reliable (posterior median =  0.014, HDI [ 0.010, 0.019], pd = 100.00%).  We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
  -The effect of beliefs in *Spirit Excludes God*  is reliably positive (posterior median =   0.018, HDI [ 0.001, 0.035], pd = 96.05%). We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.051, HDI [ 0.031, 0.071], pd = 100.00%). We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.076, HDI [ 0.048, 0.102], pd = 100.00%). We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with skeptics.

## Life Satisfaction: years X beliefs

```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
```
  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.012, HDI [ 0.007, 0.017], pd = 100.00%). We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.010, HDI [ 0.004, 0.015], pd = 99.90%). We infer that people who believe in a spirit or life force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.002, HDI [-0.011, 0.007], pd = 62.80%). We infer that people who believe in a God but not a spirit or life force tended to grow over time in their life satisfaction at a similar rate to skeptics.

@@

### Authors contributions
BH and JB conceived of the study. BH wrote the first draft of the manuscript. EW improved the approach with BH and JB and revised the manuscript. DD revised the manuscript. CS was responsible for NZAVS data collection and revised the manuscript. JB performed the data re-analysis and revised the manuscript.

@@

# Appendix 
## Model Equations
### Priors for the Personal Wellbeing model

### 2020-11-27 Supplement_And_Analysis.Rmd
close to final revision - ben to proof on overleaf.

---
title             : "Supplement: Analysis for Longitudinal Evidence for Growth in Subjective well-being from Spiritual Belief"
author: "contact: joseph.bulbulia@gmail.com on behalf of Highland et al."
date: "27/11/2020"
authornote: |
  <!-- https://tinyurl.com/ybremelq -->
# keywords          : "keywords"
wordcount         : no
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
# documentclass     : "apa6"
classoption       : "man"
output            : "html_document" 
editor_options: 
  chunk_output_type: console
---

@@

```{r latex_data_table, include=TRUE}
# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
#demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package. 
```

@@

@@

@@

The NZAVS is an ongoing longitudinal study. Our ethics requires us to protect the privacy of our participants. For this reason, we cannot post the raw data, even anonymous data. However, we provide data by request to qualified researchers for the purposes of replication. We areare interested in collaboration with qualified researchers: please contact Chris G. Sibley: [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).

## Credits
Package credits:

```{r, sessioninfo, include=FALSE}
#  custom function for citing packages. 
cite_packages()
```

>>>>

---
title             : "Supplement: Analysis for National Longitudinal Evidence for Growth in Subjective well-being from Spiritual Belief"
author: "contact: joseph.bulbulia@gmail.com on behalf of Highland et al."
date: "27/11/2020"
authornote: |
  <!-- https://tinyurl.com/ybremelq -->
# keywords          : "keywords"
wordcount         : no
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
# documentclass     : "apa6"
classoption       : "man"
output            : "html_document" 
editor_options: 
  chunk_output_type: console
---

@@

```{r latex_data_table, include=FALSE}
# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package. 
conflicts()
furniture::table1(d_3, splitby = ~Wave,
                    Age, 
                    NZdep,
                    Education,
                    Employed,
                    Ethnic_Categories,
                    Male,Has_Partner, 
                    Political_Conservativism,
                    Urban, 
                    Beliefs, 
                    LIFESAT,
                    PWI,
                    overall=F, 
                    output = "latex2", 
                    booktabs=TRUE)
```

@@

```{r simple_table, cache =TRUE}
bayestestR::describe_posterior(mod.1_b, ci = 0.90,  test = c("p_direction"),diagnostic = "ESS")
```

```{r simple_table_ls, cache =TRUE}
bayestestR::describe_posterior(mod.2_b, ci = 0.90, test = c("p_direction"), diagnostic = "ESS")
```

@@

```{r}
print(citation("brms"), bibtex=TRUE)
```

@@

The NZAVS is an ongoing longitudinal study. Our ethics requires us to protect the privacy of our participants. For this reason, we cannot post the raw data, even anonymous data. However, we provide data by request to qualified researchers for the purposes of replication. We areare interested in collaboration with qualified researchers: please contact Chris G. Sibley [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).
